Active inference on discrete state-spaces: a synthesis

### 2020-07-20 Response to reviewers 1.tex

\documentclass{article}

%%%Packages
\usepackage[utf8]{inputenc}
\usepackage{amsmath, amsthm, amssymb, amsfonts, dsfont, mathrsfs, dirtytalk, tikz-cd, adjustbox, url, nicefrac, booktabs,array,longtable, tabu,bm, hyperref, graphicx, geometry,comment}
\usepackage{caption}

\usepackage[T1]{fontenc}    
%\usepackage[margin=1in]{geometry}
\usepackage[english]{babel}
%\usepackage[pagebackref=true]{hyperref}
%\usepackage{hyperref}
\usepackage[capitalise]{cleveref}
%\usepackage{arxiv}

%%%Backreferences
%\renewcommand*{\backref}[1]{}
%\renewcommand*{\backrefalt}[4]{%
%    \ifcase #1%
%          \or (p.~#2.)%
 %         \else (p.~#2.)%
  %  \fi%
   % }

%%%%Macros
\newcommand{\e}{\epsilon}
\newcommand{\f}{\forall}
\newcommand{\R}{\mathbb R}
\newcommand{\st}{\text{ s.t. }}
\newcommand{\ex}{\exists}
\newcommand{\p}{\mathbb P}
\newcommand{\D}{\mathcal D(\Omega)}
\newcommand{\N}{\mathbb N}
\newcommand{\oo}{\Omega}
\newcommand{\harpoon}{\rightharpoondown}
\newcommand{\E}{\mathbb E}

%%%%Theorem environments%%%%
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{example}[theorem]{Example}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{remark}[theorem]{Remark}
\newenvironment{claim}[1]{\par\noindent\emph{Claim:}\space#1}{}
\newenvironment{proofclaim}[1]{\par\noindent\textit{Proof of claim.}\space#1}{\hfill $\blacksquare$}
\newenvironment{sketchproof}[1]{\par\noindent\textit{Sketch proof.}\space#1}{\hfill $\square$}

%% Use the option review to obtain double line spacing
%% \documentclass[authoryear,preprint,review,12pt]{elsarticle}

%% Use the options 1p,twocolumn; 3p; 3p,twocolumn; 5p; or 5p,twocolumn
%% for a journal layout:
%% \documentclass[final,1p,times,authoryear]{elsarticle}
%% \documentclass[final,1p,times,twocolumn,authoryear]{elsarticle}
%% \documentclass[final,3p,times,authoryear]{elsarticle}
%% \documentclass[final,3p,times,twocolumn,authoryear]{elsarticle}
%% \documentclass[final,5p,times,authoryear]{elsarticle}
%% \documentclass[final,5p,times,twocolumn,authoryear]{elsarticle}

%% For including figures, graphicx.sty has been loaded in
%% elsarticle.cls. If you prefer to use the old commands
%% please give \usepackage{epsfig}

%% The amssymb package provides various useful mathematical symbols
%\usepackage{amssymb}
%% The amsthm package provides extended theorem environments
%% \usepackage{amsthm}

%% The lineno packages adds line numbers. Start line numbering with
%% \begin{linenumbers}, end it with \end{linenumbers}. Or switch it on
%% for the whole article with \linenumbers.\usepackage{lineno}

%\journal{Journal of Mathematical Psychology}

\begin{document}
Editor and Reviewer comments:

\textbf{Reviewer #1:} This paper presents a detailed and thorough overview of the theory of active inference on discrete space (and action) spaces. It rapidly walks the reader through the standard formulation of the generative model, approximate posterior, and update rules and also covers some more advanced cases such as inferring the parameters of the generative model (the A,B, and D matrices), and bayesian structure learning of the  generative model.

Overall the paper is very well written and comprehensive and, while containing no novel results (with the possible exception of appendix B), provides a thorough and approachable overview of the field which is difficult to obtain otherwise due to the many disparate threads and model formulations in the literature. As such I think the manuscript provides a valuable contribution to the literature and should be accepted.

Beyond this I have some comments on the main text of the manuscript as well as a fair number of comments on appendix B, where I believe some novel derivations were presented.

Main text:

On page 15, it talks about the likelihood matrix (A) denoting the probabilities of states given observations. I believe this should be the other way around (observations given states). The former would correspond to the posterior.

Well spotted, we have changed this to “the probability of outcomes given states”.

The glossary of notation in pages 15-17 is very helpful. It would be very helpful to add, however, the types of variables specifically for the discrete state space formulation which is tackled in this paper. I.e. explicitly note that are represented as parameters of categorical distributions (and as such as vectors of real numbers). That the approximate posterior is a vector of real numbers, that the free energy and expected free energy are functionals that evaluate to a scalar quantity. Being explicit about this helps the reader 'type-check' the maths and can forestall much confusion.

Good point. We have added “that evaluates to a scalar quantity” for the free energy and expected free energy.

p21 - is 'mathematical yoga' the expression that was intended? I like it though.

Yes, this was on purpose. I am glad that you like it.

p22 -- it should be clarified that the VFE does not particularly measure surprise except when the approximate and true posterior are very close. Rather it is a bound on the true posterior, and since the true posterior is generally intractable, how tight this bound is is generally unknowable.

Thanks. We implemented this.

For equations 7,8,9 I think some further explanation of how to reach one from the other would be helpful here (if only in an appendix). Refreshing the reader's memory about how the probability densities (and expectations) are represented in terms of matrix operations would make this transition easier to follow. In equation 9, I don't believe $v_dot$ or v are defined anywhere (are these the prediction errors). Also using $v_dot$ is odd in a discrete time formulation. Could this just be formulated as a discrete time gradient descent - i.e. $v_{t+1} = v_t + \grad F$.

Done

Also some further clarification of how exactly equation 8 which involves essentially the difference between the approximate posterior and the generative model maps onto the concept of prediction errors as in predictive coding would be appreciated. For instance, is the posterior or the generative model meant to be considered the 'prediction'?

Ok

In equation 12, I don't believe $s_\pi_\tau$ is defined anywhere. Indeed, what does it mean to just have a state instead of a probability density here?

I don’t understand

on page 29, the pederson citation doesn't have any date associated with it in the text

Ok

in equation 17, should the sums run up to T (not t), as in they should be up to the time horizon?

\textit{No. The sum should run over the time-steps where outcomes have been observed and in the paper we consider that outcomes have been observed up to time t. But the software implementation of learning in active inference occurs once at the end of every trial, i.e., when $t=T$. We have the following paragraph that explains this and hopefully solves your concerns. "Since synaptic plasticity dynamics occur at a much slower pace than perceptual inference, it is computationally much cheaper -- in numerical simulations -- to do a one-step belief update at the end of each trial of observation epochs. Explicitly, setting the free energy gradient to zero at the end of the trial gives the following update for Dirichlet parameters:}
	\begin{equation}
	\label{eq: learning dirichlet one step}
  	    \bold a=  a +\sum_{\tau =1}^T o_\tau \otimes \bold s_\tau"
  	\end{equation}

The structure learning equations (eq 22-26) use a completely different notation than the rest of the paper. Unless there is compelling reason not to do this, I think it would be good to bring them into alignment.

\textit{We have changed all the nu's (the notation that didn't coincide) with A's (notation that coincides). In addition, we added "In the following, we show BMR for learning the likelihood matrix $A$. Note that BMR is generic and could be used on any other variable that may be optimised during learning, just by replacing $A$ in the following lines."}

on page 52, the date on the mirza paper is weird (goes into the day)

Thanks we corrected this.

on page 53 this paragraph:
Examples of
660 generative models that do not fall within the current discrete state-space, continuous
661 state-space (Buckley et al., 2017; Friston et al., 2012c; Adams et al., 2013b; Brown
662 et al., 2013; Adams et al., 2013a; Friston et al., 2012b; Brown and Friston, 2012)
663 or mixed (Friston et al., 2017c; Parr and Friston, 2018d, 2019) models - currently
664 implemented in active inference - include Markov decision trees (Jordan et al., 1998,
665 1997) and Boltzmann machines (Stone, 2019; Ackley et al., 1985; Salakhutdinov and
666 Hinton, 2012)

seems garbled to me

\textit{We have simplified the paragraph and instead have: "Examples of generative models that are used in learning and decision-making, yet are not accommodated by the current process theory, include Markov decision trees and Boltzmann machines.}

In appendix B is where novel (to me) results are presented. While the maths seems valid, I had some concerns about the presentation and explanation of the ideas. My core worry or confusion is about the justification of the Gibbs free energy. This quantity is introduced without any motivation or justification for why it should be considered. While the derivations about policies follow directly from the definition ,it is unclear that the gibbs free-energy is unique in this sense. Effectively, the core idea is to define a 'free energy' of some form G = Q(pi) - KL[Q(s,a)||p(s,a)] and thus that when the KL between the predicted and steady state is 0, then Q(pi) = G. However this specifies a very large class of possible functionals (by multiplying and dividing by different quantities) and it is unclear why the relationship between these functionals which are straightforwardly derived 'backwards' and the KL between the predicted and steady state is important at all. I think the
logic behind these manipulations should be spelled much more clearly out in the text to motivate why the gibbs free energy or some variant is useful to consider in this context.

Ok

In this sentence: "These solutions exist in virtue of conditional independencies, where the hidden states provide a Markov blanket that separates policies from outcomes". It is claimed that solutions for these policies always exist. This is not at all clear in the general case and I think needs stronger justification. It is easy to imagine cases where no possible policy allows you to reach a steady state density -- for instance if the steady state density exists on the other side of an impassable gulf. I feel like there are important smoothness (and perhaps ergodicity?) assumptions on the state and action space that are being elided here and which would be interesting to make explicit.

?

KL control is presented as being a special case of the gibbs free energy when beta = 0. However when beta = 0, that implies that the posterior density over policies is 0 (!!!) which is a very odd situation given that KL control agents can presumably be thought of as having policies.

When $\beta =0$, the posterior over policies is not zero, rather
$$Q(\pi) = \exp(-G(\pi; 0)) = \exp(-D_{KL}[Q(s_\tau, A|\pi)||P(s_\tau, A)])$$
and indeed, KL control agents do have policies.

Similarly, the expected free energy is attained when the average policy posterior is equal to the average likelihood (?) which is also a strange and nonintuitive condition.

The expected free energy corresponds to the case

I feel like a more intuitive justification and explanation of the gibbs free energy and these sorts of relations would be much appreciated in demystifying this section.

Further explanation of corollary 3 (that the markov blanket implies beta = 1 and thus the EFE is returned) would also be very much appreciated.

\textbf{Reviewer #2:} This paper summarizes the theories of active inference, a normative principle of various domains of cognition and behavior. This review covers a large body of literature regarding active inference. I feel this paper will be a useful guide for those who want to learn and active inference. I have only minor requests for better motivating readers to read this.

As this journal is the journal of mathematical *psychology*, I suggest that the authors discuss how the theory of active inference is beneficial for psychologists, who want to understand psychological or cognitive processes, without mentioning to neurophysiology. Psychologists often use mathematical models of psychological processes, e.g., drift-diffusion models and the Rescorla-Wagner model, to fit or simulate behavioral data. These mechanistic models can be easily related to psychological processes. On the other hand, it is somehow difficult to understand what computational process works in the active inference framework, especially for those who are not familiar with the concept of statistical physics. I believe that active inference offers a principle of behavior, but the authors may want to clarify these issues, e.g., how simulating experimental behavior in active inference is useful for the understanding (psychological) processes, and, how the active inference provides
a principle underlying the existing process models.

Minor comments:

1. The idea that is difficult to understand for beginners is dealing with the policy $\pi$ as a random variable. Readers may want a kind notion regarding this around equation (1).

2. Some terminology for mathematical terms is used without explaining the reason. For example, why the first term of equation (10) is termed "Risk"? --in decision science, the risk usually refers to the variance of outcome (i.e., financial risk), but this does not seem the case here. Also, this is true for "ambiguity" and "intrinsic value."

3. Line 155, "and the probability of states given outcomes": perhaps this should read "and the probability of outcomes given states"?

4.The name of the person in the photo is missing in Fig. 4 while there is in Fig.3.

\bibliographystyle{elsarticle-harv} 
\bibliography{aiexplained_bib}

\end{document}

### 2020-07-20 elsarticle-template-harv.tex
Updates from Overleaf

The generative model \citep{bishopPatternRecognitionMachine2006} expresses how the agent represents the world. This is a joint probability distribution over sensory data and the hidden (or latent) causes of these data. The sorts of discrete state-space generative models used in active inference are specifically suited to represent discrete time series and decision-making tasks. These can be expressed as variants of partially observable Markov decision processes (POMDPs; \citep{astromOptimalControlMarkov1965}): from simple Markov decision processes \citep{bartoReinforcementLearningIntroduction1992,stoneArtificialIntelligenceEngines2019,whiteMarkovDecisionProcesses2001} to generalisations in the form of deep probabilistic (hierarchical) models \citep{fristonDeepTemporalModels2018,boxMultiparameterProblemsBayesian1965,allenbyHierarchicalBayesModels2005}. For clarity, the process theory is derived for the simplest model that facilitates understanding of subsequent generalisations; namely, a POMDP where the agent holds beliefs about the probability of the initial state (specified as $D$), the transition probabilities from one state to the next (defined as matrix $B$) and the probability of states given outcomes (i.e., the likelihood matrix $A$); see Figure \ref{fig:gen mod}.

@@

$\Pi$ & Set of all allowable policies; i.e., sequences of actions. & Finite subset of $U^T$. \\\addlinespace[0.3cm]

@@

$Q$ & Approximate posterior distribution. & Probability distribution over the latent variables of the generative model.\\\addlinespace[0.3cm]

$F, F_\pi$ & Variational free energy and variational free energy conditioned upon a policy. & Functionals of $Q$.\\\addlinespace[0.3cm]

$G$ & Expected free energy. & Function defined on $\Pi$. \\\addlinespace[0.3cm]

@@

$A$ & Likelihood matrix. The probability of the state-outcome pair $o_\tau, s_\tau$ is given by $o_\tau \cdot A s_\tau$. & Random variable over the subset of $M_{n\times m}(\mathbb R)$ with columns in $\{x \in \mathbb R^{k} | x_i >0, \sum_i x_i =1\}$. \\\addlinespace[0.3cm]

$B$ & Matrix of transition probabilities from one state to the next state given action $\pi_{\tau-1}$. The probability of possible state $s_\tau$, given $s_{\tau-1}$ and action $\pi_{\tau-1}$ is $s_\tau\cdot B_{\pi_{\tau-1}}s_{\tau -1}$. & Matrix in $M_{m\times m}(\mathbb R)$ with columns in $\{x \in \mathbb R^{k} | x_i >0, \sum_i x_i =1\}$. \\\addlinespace[0.3cm]

@@

This paper will be largely concerned with deriving and interpreting the inferential dynamics that agents might implement using the generative model in Figure \ref{fig:gen mod}. We leave the discussion of more complex models to Appendix \ref{appendix: more complex models}, since the derivations are analogous in those cases.

@@

Variational Bayesian inference rests upon minimisation of a quantity called (variational) free energy, which measures the improbability (i.e., the surprise) of sensory observations, under a generative model. Simultaneously, variational free energy minimisation is a statistical inference technique that enables the approximation of the posterior distribution in Bayes rule. In machine learning, this is known as variational Bayes \citep{bishopPatternRecognitionMachine2006,jordanIntroductionVariationalMethods1998,bealVariationalAlgorithmsApproximate2003,bleiVariationalInferenceReview2017}.
Active inference agents minimise variational free energy, enabling concomitant maximisation of their model evidence and inference of the latent variables of their generative model. In the following, we consider a particular time point to be given $t\in\{1,...,T\}$, whence the agent has observed a sequence of outcomes $o_{1:t}$. The posterior about the latent causes of sensory data is given by Bayes rule:

@@

Computing the posterior distribution requires computing the model evidence $P(o_{1:t})=\sum_{\pi \in \Pi }\sum_{s_{1:T}\in S^T}\int P(o_{1:t} ,s_{1:T}, A, \pi ) \: dA$, which is intractable for complex generative models embodied by biological and artificial systems \citep{fristonHierarchicalModelsBrain2008} -- a well-known problem in Bayesian statistics.
An alternative to computing the exact posterior distribution is to optimise an approximate posterior distribution over latent causes $Q(s_{1:T}, A, \pi)$, by minimising the Kullback-Leibler (KL) divergence \citep{kullbackInformationSufficiency1951} ($D_{KL}$) -- a non-negative measure of discrepancy between probability distributions. We can use the definition of the KL divergence and Bayes rule to arrive at the variational free energy $F$, which is a functional of approximate posterior beliefs:

  	\begin{equation}
  	\label{eq: vfe}
  	    \begin{split}
  	        0 &\leq D_{KL}[Q(s_{1:T}, A, \pi)||P(s_{1:T}, A, \pi |o_{1:t})] \\
  	        &= \E_{Q(s_{1:T}, A, \pi)}[\log Q(s_{1:T}, A, \pi) - \log P(s_{1:T}, A, \pi |o_{1:t})] \\
  	        &= \E_{Q(s_{1:T}, A, \pi)}[\log Q(s_{1:T}, A, \pi) - \log P(o_{1:t},s_{1:T}, A, \pi) + \log P(o_{1:t})] \\
  	        &= \underbrace{\E_{Q(s_{1:T}, A, \pi)}[\log Q(s_{1:T}, A, \pi) - \log P(o_{1:t},s_{1:T}, A, \pi)]}_{F[Q(s_{1:T}, A, \pi)]} + \log P(o_{1:t}) \\
  	        \Rightarrow & -\log P(o_{1:t}) \leq F[Q(s_{1:T}, A, \pi)]
  	    \end{split}
  	\end{equation}

From \eqref{eq: vfe}, one can see that by varying $Q$ to minimise the variational free energy enables us to approximate the true posterior, while simultaneously ensuring that surprise remains low. This means that variational free energy minimising agents, simultaneously, infer the latent causes of their observations and maximise the evidence for their generative model.
To aid intuition, the variational free energy can be rearranged into complexity and accuracy:

@@

The goal is now to minimise variational free energy with respect to $Q$. To obtain a tractable expression for the variational free energy, we need to assume a certain simplifying factorisation of the approximate posterior. There are many possible forms \citep{parrNeuronalMessagePassing2019,yedidiaConstructingFreeEnergyApproximations2005,heskesConvexityArgumentsEfficient2006} (e.g., mean-field, marginal, Bethe), each of which trades off the quality of the inferences with the complexity of the computations involved. For the purpose of this paper we use a particular (structured) mean-field approximation, see Figure \ref{fig:gen mod}:

@@

In active inference, perception is equated with state estimation \citep{fristonActiveInferenceProcess2017} (e.g., inferring the temperature from the sensation of warmth), consistent with the idea that perceptions are hypotheses \citep{gregoryPerceptionsHypotheses1980}. To infer the (past, present and future) states of the environment, an agent must minimise the variational free energy with respect to $Q(s_{1:T}|\pi)$ for each policy $\pi$. This provides the agent’s inference over hidden states, contingent upon pursuing a given policy. Since the only part of the free energy that depends on $Q(s_{1:T}|\pi)$ is $F_\pi$, the agent must simply minimise $F_\pi$. 
Substituting $Q(s_{\tau}|\pi)$ by their sufficient statistics (i.e., parameters $\bold s_{\pi \tau}$), $F_\pi$ becomes a function of those parameters. This enables us to rewrite \eqref{eq: VFE_pi}, conveniently in matrix form:

\begin{equation}
    \begin{split}
        F_\pi(\bold s_{\pi 1},...,\bold s_{\pi T}) &= \sum_{\tau = 1}^T \bold s_{\pi \tau } \cdot \log \bold s_{\pi \tau } 
        - \sum_{\tau = 1}^t o_\tau \cdot \textbf{log} \bold A \bold s_{\pi \tau }\\
        &- \bold s_{\pi 1 } \log D- \sum_{\tau = 2}^T \bold s_{\pi \tau } \cdot \log (B_{\pi_{\tau-1}})\bold s_{\pi \tau-1 }
    \end{split}
\end{equation}

This enables to compute the variational free energy gradients \citep{petersenMatrixCookbook}:
\begin{equation}
    \label{eq:vfe gradients perception}
    \begin{split}
        \nabla_{\bold s_{\pi\tau}} F_\pi(\bold s_{\pi 1},...,\bold s_{\pi T})& = \vec {1} +\log  \bold s_{\pi\tau}  \\
        &- \begin{cases}
             o_\tau \cdot \textbf{log} \bold A+ \bold s_{\pi\tau+1} \cdot \log (B_{\pi_{\tau}}) + \log D \quad  \text{if } \tau = 1 \\
            o_\tau \cdot \textbf{log} \bold A+ \bold s_{\pi\tau+1} \cdot \log (B_{\pi_{\tau}}) +\log (B_{\pi_{\tau-1}}) \bold s_{\pi\tau-1} \quad \text{if } 1<\tau \leq t  \\
             \bold s_{\pi\tau+1} \cdot \log (B_{\pi_{\tau}}) +\log (B_{\pi_{\tau-1}}) \bold s_{\pi\tau-1} \quad \text{if }  \tau > t
        \end{cases}
    \end{split}
\end{equation}

The neuronal dynamics are given by a gradient descent on free energy \citep{fristonActiveInferenceProcess2017}, with state-estimation expressed as a softmax function of accumulated (negative) free energy gradients. The constant term $\vec{1}$ is generally omitted since the softmax function removes it anyway. This enables us to equate the gradient with a prediction error.

  	\begin{equation}
  	\label{eq: perceptual dynamics}
  	    \begin{split}
  	        \dot v (\bold s_{\pi 1},...,\bold s_{\pi T})&=-\nabla_{\bold s_{\pi\tau}} F_\pi(\bold s_{\pi 1},...,\bold s_{\pi T}) \\
  	        \bold s_{\pi\tau} = \sigma (v)
  	    \end{split}
  	\end{equation}

The softmax function – a generalisation of the sigmoid to vector inputs – is a natural choice as the variational free energy gradient is a logarithm and the components of $\bold s_{\pi\tau}$ must sum to one.

\subsection{Plausibility of neuronal dynamics}

@@

Furthermore, these dynamics are consistent with predictive processing \citep{raoPredictiveCodingVisual1999,bastosCanonicalMicrocircuitsPredictive2012} -- since active inference prescribes dynamics that minimise prediction error -- although they generalise it to a wide range of generative models. Note that, while also a variational free energy gradient, this sort of prediction error is not the same as that given by predictive coding schemes (which rely upon a different sort of generative model) \citep{buckleyFreeEnergyPrinciple2017,fristonVariationalFreeEnergy2007,bogaczTutorialFreeenergyFramework2017}.

@@

\begin{equation}
\label{eq: action selection}
    u_t = \arg \max_{u \in U } \left (\sum_{\pi \in \Pi} \delta_{u, \pi_t} Q(\pi)\right)
\end{equation}

where $\delta$ is the Kronecker delta. In addition, we obtain a policy independent state-estimation at any time point $\tau \in \{1,...,T\}$, as a Bayesian model average of approximate posterior beliefs about hidden states under policies:

>>>>

The generative model \citep{bishopPatternRecognitionMachine2006} expresses how the agent represents the world. This is a joint probability distribution over sensory data and the hidden (or latent) causes of these data. The sorts of discrete state-space generative models used in active inference are specifically suited to represent discrete time series and decision-making tasks. These can be expressed as variants of partially observable Markov decision processes (POMDPs; \citep{astromOptimalControlMarkov1965}): from simple Markov decision processes \citep{bartoReinforcementLearningIntroduction1992,stoneArtificialIntelligenceEngines2019,whiteMarkovDecisionProcesses2001} to generalisations in the form of deep probabilistic (hierarchical) models \citep{fristonDeepTemporalModels2018,boxMultiparameterProblemsBayesian1965,allenbyHierarchicalBayesModels2005}. For clarity, the process theory is derived for the simplest model that facilitates understanding of subsequent generalisations; namely, a POMDP where the agent holds beliefs about the probability of the initial state (specified as $D$), the transition probabilities from one state to the next (defined as matrix $B$) and the probability of outcomes given states (i.e., the likelihood matrix $A$); see Figure \ref{fig:gen mod}.

@@

$\Pi$ & Set of all allowable policies; i.e., sequences of actions. & Finite subset of $\underbrace{U \times \ldots \times U}_{T \text{ times}} =U^T$. \\\addlinespace[0.3cm]

@@

$Q$ & Approximate posterior distribution, over the latent variables of the generative model, e.g., $s_{1:T},A$. & Probability distribution. $Q(s_{\tau})$ (resp. $Q(A)$) is a categorical distribution over $S$ (resp. a product of Dirichlet distributions over the columns of $A$), which can be represented as a vector of real numbers (resp. a matrix in $M_{n\times m}(\mathbb R_{>0})$) corresponding to its parameters. \\\addlinespace[0.3cm]

$F, F_\pi$ & Variational free energy and variational free energy conditioned upon a policy. & Functionals of $Q$ that evaluate to a scalar quantity.\\\addlinespace[0.3cm]

$G$ & Expected free energy. & Function defined on $\Pi$ that evaluates to a scalar quantity. \\\addlinespace[0.3cm]

@@

$A$ & Likelihood matrix. The probability of the state-outcome pair $o_\tau, s_\tau$, namely $P(o_\tau|s_\tau,A)$ is given by $o_\tau \cdot A s_\tau$. & Random variable over the subset of $M_{n\times m}(\mathbb R)$ with columns in $\{x \in \mathbb R^{n} | x_i >0, \sum_i x_i =1\}$. \\\addlinespace[0.3cm]

$B_{\pi_{\tau-1}}$ & Matrix of transition probabilities from one state to the next state given action $\pi_{\tau-1}$. The probability of possible state $s_\tau$, given $s_{\tau-1}$ and action $\pi_{\tau-1}$ is $s_\tau\cdot B_{\pi_{\tau-1}}s_{\tau -1}$. & Matrix in $M_{m\times m}(\mathbb R)$ with columns in $\{x \in \mathbb R^{m} | x_i >0, \sum_i x_i =1\}$. \\\addlinespace[0.3cm]

$D$ & Vector of probabilities of initial state. The probability of the i-th possible state occurring at time $1$ is $D_i$. & Vector of probabilities in $\{x \in \mathbb R^{m} | x_i >0, \sum_i x_i =1\}$. \\\addlinespace[0.3cm]

@@

This paper will be largely concerned with deriving and interpreting the inferential dynamics that agents might implement using the generative model in Figure \ref{fig:gen mod}. We leave the discussion of more complex models to \ref{appendix: more complex models}, since the derivations are analogous in those cases.

@@

Variational Bayesian inference rests upon minimisation of a quantity called (variational) free energy, which bounds the improbability (i.e., the surprise) of sensory observations, under a generative model. Simultaneously, free energy minimisation is a statistical inference technique that enables the approximation of the posterior distribution in Bayes rule. In machine learning, this is known as variational Bayes \citep{bishopPatternRecognitionMachine2006,jordanIntroductionVariationalMethods1998,bealVariationalAlgorithmsApproximate2003,bleiVariationalInferenceReview2017}. Active inference agents minimise variational free energy, enabling concomitant maximisation of their model evidence and inference of the latent variables of their generative model. In the following, we consider a particular time point to be given $t\in\{1,...,T\}$, whence the agent has observed a sequence of outcomes $o_{1:t}$. The posterior about the latent causes of sensory data is given by Bayes rule:

@@

Computing the posterior distribution requires computing the model evidence $P(o_{1:t})=\sum_{\pi \in \Pi }\sum_{s_{1:T}\in S^T}\int P(o_{1:t} ,s_{1:T}, A, \pi ) \: dA$, which is intractable for complex generative models embodied by biological and artificial systems \citep{fristonHierarchicalModelsBrain2008} -- a well-known problem in Bayesian statistics.
An alternative to computing the exact posterior distribution is to optimise an approximate posterior distribution over latent causes $Q(s_{1:T}, A, \pi)$, by minimising the Kullback-Leibler (KL) divergence \citep{kullbackInformationSufficiency1951} $D_{KL}$ -- a non-negative measure of discrepancy between probability distributions. We can use the definition of the KL divergence and Bayes rule to arrive at the variational free energy $F$, which is a functional of approximate posterior beliefs:

  	\begin{equation}
  	\label{eq: vfe}
  	    \begin{split}
  	        0 &\leq D_{KL}[Q(s_{1:T}, A, \pi)||P(s_{1:T}, A, \pi |o_{1:t})] \\
  	        &= \E_{Q(s_{1:T}, A, \pi)}[\log Q(s_{1:T}, A, \pi) - \log P(s_{1:T}, A, \pi |o_{1:t})] \\
  	        &= \E_{Q(s_{1:T}, A, \pi)}[\log Q(s_{1:T}, A, \pi) - \log P(o_{1:t},s_{1:T}, A, \pi) + \log P(o_{1:t})] \\
  	        &= \underbrace{\E_{Q(s_{1:T}, A, \pi)}[\log Q(s_{1:T}, A, \pi) - \log P(o_{1:t},s_{1:T}, A, \pi)]}_{=:F[Q(s_{1:T}, A, \pi)]} + \log P(o_{1:t}) \\
  	        \Rightarrow & -\log P(o_{1:t}) \leq F[Q(s_{1:T}, A, \pi)]
  	    \end{split}
  	\end{equation}

From \eqref{eq: vfe}, one can see that by varying $Q$ to minimise the variational free energy enables us to approximate the true posterior, while simultaneously ensuring that surprise remains low. The former offers the intuitive interpretation of the free energy as a generalised prediction error. Indeed, minimising free energy corresponds to suppressing the discrepancy between predictions, i.e., $Q$, and the actual state of affairs, i.e., the posterior. Altogether, this means that variational free energy minimising agents, simultaneously, infer the latent causes of their observations and maximise the evidence for their generative model. One should note that the free energy equals the surprise $-\log P(o_{1:t})$ only at the global free energy minimum, when the approximate posterior $Q(s_{1:T}, A, \pi)$ equals to the true posterior $P(s_{1:T}, A, \pi |o_{1:t})$. Outside of the global free energy minimum, the free energy upper bounds the surprise, in which case, since the true posterior is generally intractable, the tightness of the bound is generally unknowable.

To aid intuition, the variational free energy can be rearranged into complexity and accuracy:

@@

The goal is now to minimise variational free energy with respect to $Q$. To obtain a tractable expression for the variational free energy, we need to assume a certain simplifying factorisation of the approximate posterior. There are many possible forms (e.g., mean-field, marginal, Bethe, see \citep{parrNeuronalMessagePassing2019,yedidiaConstructingFreeEnergyApproximations2005,heskesConvexityArgumentsEfficient2006}), each of which trades off the quality of the inferences with the complexity of the computations involved. For the purpose of this paper we use a particular structured mean-field approximation, see Figure \ref{fig:gen mod}:

@@

In active inference, perception is equated with state estimation \citep{fristonActiveInferenceProcess2017} (e.g., inferring the temperature from the sensation of warmth), consistent with the idea that perceptions are hypotheses \citep{gregoryPerceptionsHypotheses1980}. To infer the (past, present and future) states of the environment, an agent must minimise the variational free energy with respect to $Q(s_{1:T}|\pi)$ for each policy $\pi$. This provides the agent’s inference over hidden states, contingent upon pursuing a given policy. Since the only part of the free energy that depends on $Q(s_{1:T}|\pi)$ is $F_\pi$, the agent must simply minimise $F_\pi$. 
Substituting $Q(s_{\tau}|\pi)$ by their sufficient statistics (i.e., parameters $\bold s_{\pi \tau}$), $F_\pi$ becomes a function of those parameters. This enables us to rewrite \eqref{eq: VFE_pi}, conveniently in matrix form (see \ref{app: deriving perceptual dynamics} for details):

\begin{equation}
\label{eq: vfe pi matrix form perception}
  	    \begin{split}
        F_\pi(\bold s_{\pi 1},...,\bold s_{\pi T}) &= \sum_{\tau = 1}^T \bold s_{\pi \tau } \cdot \log \bold s_{\pi \tau } 
        - \sum_{\tau = 1}^t o_\tau \cdot \textbf{log} \bold A \bold s_{\pi \tau }\\
        &- \bold s_{\pi 1 } \log D- \sum_{\tau = 2}^T \bold s_{\pi \tau } \cdot \log (B_{\pi_{\tau-1}})\bold s_{\pi \tau-1 }
    \end{split}
\end{equation}

This enables to compute the variational free energy gradients \citep{petersenMatrixCookbook2012}:
\begin{equation}
    \label{eq:vfe gradients perception}
    \begin{split}
        \nabla_{\bold s_{\pi\tau}} F_\pi(\bold s_{\pi 1},...,\bold s_{\pi T})& = \vec {1} +\log  \bold s_{\pi\tau}  \\
        &- \begin{cases}
             o_\tau \cdot \textbf{log} \bold A+ \bold s_{\pi\tau+1} \cdot \log (B_{\pi_{\tau}}) + \log D \quad  \text{if } \tau = 1 \\
            o_\tau \cdot \textbf{log} \bold A+ \bold s_{\pi\tau+1} \cdot \log (B_{\pi_{\tau}}) +\log (B_{\pi_{\tau-1}}) \bold s_{\pi\tau-1} \quad \text{if } 1<\tau \leq t  \\
             \bold s_{\pi\tau+1} \cdot \log (B_{\pi_{\tau}}) +\log (B_{\pi_{\tau-1}}) \bold s_{\pi\tau-1} \quad \text{if }  \tau > t
        \end{cases}
    \end{split}
\end{equation}

The neuronal dynamics are given by a gradient descent on free energy \citep{fristonActiveInferenceProcess2017}, with state-estimation expressed as a softmax function of accumulated (negative) free energy gradients, that we denote by $v_{\pi \tau}$ (see Section \ref{sec: plausibility of neuronal dynamics} for an interpretation). The constant term $\vec{1}$ is generally omitted since the softmax function removes it anyway.

  	\begin{equation}
  	\label{eq: perceptual dynamics}
  	    \begin{split}
  	        \dot v_{\pi \tau} (\bold s_{\pi 1},...,\bold s_{\pi T})&=-\nabla_{\bold s_{\pi\tau}} F_\pi(\bold s_{\pi 1},...,\bold s_{\pi T}) \\
  	        \bold s_{\pi\tau} = \sigma (v_{\pi \tau})
  	    \end{split}
  	\end{equation}

The softmax function $\sigma$ – a generalisation of the sigmoid to vector inputs – is a natural choice as the variational free energy gradient is a logarithm and the components of $\bold s_{\pi\tau}$ must sum to one. Note that although we have expressed the inferential dynamics as a gradient descent in continuous time, any numerical implementation of active inference would implement a discretised version of \eqref{eq: perceptual dynamics} until convergence, for example
  	\begin{equation*}
  	    \begin{split}
  	         v_{\pi \tau}^{(k)}&= v_{\pi \tau}^{(k-1)} -\kappa \nabla_{\bold s^{(k-1)}_{\pi\tau}} F_\pi(\bold s^{(k-1)}_{\pi 1},...,\bold s^{(k-1)}_{\pi T}) \text{ for small } \kappa >0 \\
  	        \bold s_{\pi\tau}^{(k)} &= \sigma (v_{\pi \tau}^{(k)}).
  	    \end{split}
  	\end{equation*}

\subsection{Plausibility of neuronal dynamics}
\label{sec: plausibility of neuronal dynamics}

@@

Furthermore, these dynamics are consistent with predictive processing \citep{raoPredictiveCodingVisual1999,bastosCanonicalMicrocircuitsPredictive2012} -- since active inference prescribes dynamics that minimise prediction error -- although they generalise it to a wide range of generative models. Note that, while also a variational free energy gradient, this sort of prediction error is not the same as that given by predictive coding schemes (which rely upon a different sort of generative model, see \citep{buckleyFreeEnergyPrinciple2017,fristonVariationalFreeEnergy2007,bogaczTutorialFreeenergyFramework2017}).

@@

\begin{equation}
\label{eq: action selection}
    u_t = \arg \max_{u \in U } \left (\sum_{\pi \in \Pi , \pi_t =u} Q(\pi)\right).
\end{equation}
